A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-02-24 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: resumed working on the article

\def\>{\rangle}
\def\<{\langle}
\DeclareMathOperator{\erf}{erf}

@@

The goal of this paper is to show that, in the same way that classical probability can be axiomatized through standard measure theory, quantum probability, and in fact any probability used a physical theory, can be axiomatized through non-additive (i.e. fuzzy) measure theory. We will define a generalized concept of probability that is ultimately based on physically motivated axioms, giving the physicist a clear operational motivation to our definitions, the mathematicians a formally well-defined mathematical structure and the philosopher a unified framework in which probability can be interpreted in the context of physical theories. We are not going to be able to develop a full theory here, as there are still a number of conceptual and technical challenges to be solved that will require work in physics, mathematics and philosophy. However, we will show that there are enough results that point in a good direction, and therefore make it reasonable to invest time and effort to work on the open problems. For those interested, the current state of the research will always be available in an open document that will be routinely updated as the theory matures.

@@

@@

@@

As mentioned in the introduction, our goal is to construct a probability theory of general applicability to physical systems. Our approach is not to define an axiomatic theory of probability and apply it to physical systems, but rather to characterize basic properties of physical states and from those build a theory of probability that fits those definitions. As is common in our larger project, we start from axioms that, in our opinion, represent requirements of scientific practice. The advantage is that the realm of applicability and physical interpretation will be clear from the start.

We start from requiring a physical theory to allow statistical descriptions: at the very least, it will provide us with a notion of ensemble, and the space of ensembles that are allowed by the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions on phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will in principle be different, they will all share common features: they must allow measurable properties of the system, probability distributions for outcomes of said properties, characterize reversible and irreversible processes through an entropy function, and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must allow statistical descriptions, i.e.~provide an ensemble space. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterize experimentally. The approach, then, can most likely be given firm conceptual grounds, which we leave as an open but solvable philosophical problem.

@@

\subsection{Experimental verifiability}
The first requirement is that ensembles must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Based on previous work, this mathematically requires an ensemble space to be a $T_0$ second-countable topological space, where open sets represent verifiable statements and Borel sets represent statements associated with a test, but without guarantee of termination.

@@

Ultimately, the convex structure is responsible for all linear structures we have in physics. Both in classical and quantum mechanics, the ensemble space is not just a convex set, but a convex subset of a vector space. In those cases, in fact, the convex structure is, in a sense, invertible. If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. The ensemble space is \textbf{complemented}, meaning that there is only one complement to a component of an ensemble.\footnote{In the context of convex spaces, this property is called ``cancellative.''}

This perspective allows us to understand what negative probability is. Suppose we have a real vector space of dimensions $n$, then $n$ linearly independent vectors are enough to express every point as a linear combination. If we add another point, we can now express every point as an affine combination, that is a linear combination where the coefficients sum to one, though some can still be negative. Since the space of mixed states in quantum mechanics is a real vector space, we can choose a linearly dependent set of states that span the whole space, and every mixed state can be expressed as an affine combination, as a pseudo-probability.

Pseudo-probability distributions, then, require the additional vector space structure. However, we found no physical justification to require the additional structure, and we suspect that it may need to be relaxed in future physical theories.  Therefore it is important that our fundamental definitions require only the convex structure, and not the vector space structure.

@@

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined \textbf{entropy}. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following:
\begin{enumerate}
	\item strictly concave: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item bounded from above: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{It can be shown that if an upper bound that only depends on the mixing coefficients exists, then it must be the Shannon entropy.}

@@

Another important feature is that if two ensembles saturate the upper bound, then they are orthogonal in both classical and quantum mechanics. Therefore we define two ensembles to be \textbf{orthogonal} if their mixture leads to a maximal entropy increase. An additional axiom is valid for orthogonal ensembles. We say an ensemble is a component of another ensemble if the second can be expressed as a mixture of the first with some other ensemble. On physical grounds, two orthogonal ensembles $\ens_1, \ens_2 \in \Ens$ are perfectly distinguishable and must have no common component. That is, there is no $\ens_3 \in \Ens$ that is a component of both $\ens_1$ and $\ens_2$. Note that the converse is not generally true: two pure states in quantum mechanics have no components in common, but are not necessarily orthogonal. We find that an ensemble space where no common component implies orthogonality can be described by a space of classical probability measures. Non-classicality, then, is exactly the failure of this property.

\begin{tcolorbox}[colback=white, colframe=black]
	Existence of entropy $\Rightarrow$ Strictly concave function whose upper bound defines orthogonality.
\end{tcolorbox}

@@

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. However, some infinite mixtures lead to physically ill-defined systems, such as a particle with an infinite average position. It is the connection to experimental verification, the topology, that will tell us which limits are allowed and which are not. Conceptual clarity on these details is essential to find the correct generalization. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensembles perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i=1}^{6} p_i \ens_i = 1$ of those pure states $\ens_i$. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work and there is a temptation to think this is how everything works. For example, in classical mechanics the pure states would be perfect preparations of position and momentum, and ensembles are probability distributions. Not only is this problematic mathematically, it is conceptually not sustainable.

On physical grounds, we cannot consistently prepare classical particles perfectly, therefore there is no ensemble corresponding to a perfect preparation of a point of phase space. The mathematical facts that delta Dirac are not proper functions, their support has an empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. One may concede that the ensembles do not exist physically, but may still want them to consider them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points are the limit of this decomposition. Probability distributions are defined on these limits. This kind of works in classical mechanics, but in general.

In quantum mechanics we do have pure states, and these are ensembles that cannot be further decomposed into finer ensembles. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure. It is the limit where we are preparing a pure state on a smaller and smaller region of space. It is the limit of a sequence of pure states, a limit in the topology. Physically, more than mathematically, it is a completely different thing. It is only when we realized this difference that a clean route to generalization presented itself.\footnote{The fact that pure states can be seen, in some sense, as superposition of eigenstates of position, makes the matter even more confusing.}

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide in classical discrete spaces, which are the least physically interesting case as they do not even allow classical particle mechanics. Equivalently, \textbf{mixtures are about preparations, not measurements, and probability distributions are about measurements, not preparations}. Since they can only coincide in classical spaces, a generalized notion of probability for physical theories must resolve and characterize the distinction.

To recap, the topological structure captures experimental verifiability and is responsible for the limits and the connection to measureable quantities. The convex structure captures statistical mixing and is responsible for all linear structures of physical theories. The entropic structure captures the variability within an ensemble and is ultimately responsible for all geometric structure. We now concentrate on recovering the measure theoretic structures associated to probability and state counting.

@@

Note that the mixing coefficient $p$ is connected to probability but it is not itself probability in the standard sense. It represents a parameter for the mixing states, mixing preparations, while probability is typically understood as the chance of something happening, the likelihood of a final event. Mathematically, the probability measure is a function from an event, which can be understood as a collection of fully distinct outcomes, to a number from zero to one. If we are mixing ensembles that represent fully distinct outcomes (i.e. they are orthogonal), then the mixing $p$ is a probability in the standard sense. If we are mixing ensembles that do not represent fully distinct outcomes, that are not orthogonal like spin up and spin left, the probability of getting one of the outcomes is not the mixing coefficient $p$.

We essentially have two ways in which two events can be compared. We have \textbf{disjoint events}, those that represent disjoint sets of different states, like spin up and spin down; we also have \textbf{orthogonal events}, those that represent represent orthogonal sets of states, like spin up and spin down. The standard notion of probability applies only when considering orthogonal events. Classical probability rests on the assumption that all disjoint events are orthogonal. That is, that all states can be perfectly distinguished. Quantum mechanics breaks this assumption, and therefore we need a probability theory that can deal with disjoint but non-orthogonal events.

Let us concentrate on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definition.

@@

Note that a probability satisfies all the properties of a fraction capacity, but it also satisfy on additional property: additivity over disjoint sets. That is, $\mu(A \cup B) = \mu(A) + \mu(B)$ if $A \cup B = \emptyset$. This is not true in general, as we can see in a simple example from quantum mechanics. Let $\Ens$ be the ensemble space of a qubit, that is the space of density matrices of a two dimensional Hilbert space. Geometrically, this is the convex set represented by a three dimensional ball. Let $\ens_{\psi} = |\psi\>\<\psi|$ be a pure state and let $\ens_{\phi} = |\phi\>\<\phi|$ be its orthogonal. Geometrically, these are two opposite points on the surface of the ball. The maximally mixed state can be expressed as $\ens = \frac{1}{2} \ens_0 + \frac{1}{2} \ens_1$. This is the center point of the ball and is in fact the midpoint between two opposite points. Since this is valid for any point, we have $\frcap_{\ens}(\{\ens_{\psi}\})=\frac{1}{2}$ for any $\ens_{\psi} \in \Ens$ and $\frcap_{\ens}(\{\ens_{\psi}, \ens_{\phi}\})=1$ for any pair of orthogonal pure states. If $A$ is a set of two orthogonal pure states, and $B$ is state different from the two, we have $\frcap_{\ens}(A \cup B) = 1 < \frcap_{\ens}(A) + \frcap_{\ens}(B) = \frac{3}{2}$.

>>>>

\def\ortho{\perp}
\def\northo{\nperp}
\def\separate{\downmodels}
\def\nseparate{\ndownmodels}

\def\>{\rangle}
\def\<{\langle}

@@

The goal of this paper is to show that, in the same way that classical probability can be axiomatized through standard measure theory, quantum probability, and in fact any probability used a physical theory, can be axiomatized through non-additive (i.e. fuzzy) measure theory. We will define a generalized concept of probability that is ultimately based on physically motivated axioms, giving the physicist a clear operational motivation to our definitions, the mathematicians a formally well-defined mathematical structure and the philosopher a unified framework in which probability can be interpreted in the context of all physical theories. We are not going to be able to develop a full theory here, as there are still a number of conceptual and technical challenges to be solved that will require work in physics, mathematics and philosophy. However, we will show that there are enough results that point in the same direction, and therefore make it reasonable to invest time and effort to work on the open problems.

The material presented is part of a larger work that aims to coalesce ideas and results from different branch of mathematics and physics. For those interested, the current state of the research is openly available in an open document that will be routinely updated as experts from different fields review and help us improve the overall theory.

@@

Similarly, given a set $A$ of ensembles we define the \textbf{state capacity} as the exponential of the highest entropy obtainable with a mixture of $A$. Note that the highest entropy of a subspace in both classical and quantum statistical mechanics is given by the uniform distribution, therefore the state capacity recovers the count of states (i.e. the Liouville volume) in classical mechanics and the dimensionality of the Hilbert space in quantum mechanics. The state capacity is also a non-negative, sub-additive and continuous set function. This points to a possible generalization of measure theoretic calculus to a sub-additive case that would be guaranteed to work on all physical theories.

@@

The standard way to characterize probability is through measure theory. Given a sample space $\Omega$, which represents the possible outcomes, and a $\sigma$-algebra $\Sigma_{\Omega}$, which represents the events, we define a probability measure $p : \Sigma_{\Omega} \to [0,1]$. Note that the probability is assigned to events, not the outcomes. This is what allows to treat discrete and continuous spaces within the same framework. In physics, all sample spaces are ultimately topological spaces, and the $\sigma$-algebra is typically the Borel algebra (i.e. the smallest $\sigma$-algebra that contains all open sets).

In classical mechanics, the sample space is phase space $M$ given by all possible values of position and momentum for all particles under considerations. Phase space is equipped with another measure, the Liouville measure $\mu : \Sigma_M \to [0, +\infty]$, that returns volume of each region. In statistical mechanics, this volume represents the count of states and is another key ingredient. The probability density over states is given by the Radon-Nikodym derivative $\rho = \frac{dp}{d\mu}$ and the entropy is given by $S(\rho) = - \int_M \rho \log \rho d\mu$. In case of a uniform distribution $\rho_U$ with support $U$, the entropy is given by $S(\rho_U) = \log \mu(U)$.\footnote{Note that entropy, the count of states and the symplectic form that characterizes phase space are deeply intertwined, yet they require different mathematical framework to express them. We suspect that they are "shadows" of a more fundamental mathematical structure, which our work on ensembles space is trying to identify.}

The above definitions do not apply to quantum mechanics. In the Hilbert space formulation, mixed states are represented by density operator (i.e. positive semi-definite self-adjoint operators with trace one). Note that if $\rho(x)$ is an integrable function such that $\int_X \rho(x) dx = 1$, the map $\psi(x) \mapsto \rho(x) \psi(x)$ is a density operator. Therefore the space of probability densities over position are mixed states in quantum mechanics. However, no joint distribution between position and momentum can be given, therefore quantum states cannot be described by the space of probability measures over a single sample space.

Alternative representations over phase space do exist, and they are widely used.(CITE) The use quasi-probability measure, meaning that the measure over the full space is still unitary (i.e. $\mu(M) =1$). However, the measure can be negative for smaller regions, which means it can also be grater than $1$, and there is no simple characterization of which quasi-probability measures are allowed and which aren't. Moreover, there is no simple link between then entropy of the distribution and the Liouville measure.

@@

Note that our goal is not to construct an abstract probability theory that applies to everything, but rather to construct a specific probability theory that is guaranteed to apply to all physical systems. Therefore we do not to define an axiomatic theory of probability and apply it to physical systems, but rather characterize basic properties of physical states and from those build a theory of probability that fits those definitions. Whether it is applicable to decision theory, economics, or psychology is out of scope. As is common in our larger project, we start from axioms that, in our opinion, represent requirements of scientific practice. We call these \textbf{constitutive assumptions} as they are essential to proceed with the scientific endeavor. The advantage is that the realm of applicability and physical interpretation will be clear from the start.

We start from requiring a physical theory to allow statistical descriptions: at the very least, it will provide a notion of ensemble and describe which ensembles that are allowed by the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of probability distributions on phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will in principle be different, they will all share common features: they must allow measurable properties of the system, probability distributions for outcomes of said properties, characterize reversible and irreversible processes through an entropy function, and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must allow statistical descriptions, i.e.~provide an ensemble space. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterize experimentally. Moreover, repeatability requires the ability to always test ``one more time,'' which implies that ensembles are infinite collections, and therefore idealization. The approach, then, can most likely be given firm conceptual grounds, which we leave as an open but likely solvable philosophical problem.

@@

\subsection{Experimental verifiability}
The first requirement is that ensembles must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Based on previous work, (CITE) this mathematically requires an ensemble space to be a $T_0$ second-countable topological space, where open sets represent verifiable statements and Borel sets represent statements associated with a test, but without guarantee of termination.

@@

Ultimately, the convex structure is responsible for all linear structures we have in physics. In face, the entropy and its bounds, defined later, forces the ensemble space to be the subset of a convex subset of a vector space. That is, the convex structure is ``invertible.''\footnote{If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. In the context of convex spaces, this property is called ``cancellative.''} What is still not clear is whether the ensemble space embeds continuously in a topological vector space. While topological vector spaces are well established in the literature, topological convex spaces are not.

Regardless of the topology, we are already equipped to understand what negative probability is. Suppose we have a real vector space of dimensions $n$, then $n$ linearly independent vectors are enough to express every point as a linear combination. We can also take an additional vector and express every point as an affine combination, that is a linear combination where the coefficients sum to one, though some can still be negative. Since every ensemble space, including the space of mixed states in quantum mechanics, is a subset of a real vector space, we can choose a linearly dependent set of states that span the whole space, and every mixed state can be expressed as an affine combination, as a pseudo-probability. However, the boundaries of the ensemble space will determine which affine combinations are valid states (i.e. they end up within the ensemble space), and there is no requirement that these boundary are conveniently expressible in terms of affine combinations.

The convex structure, however, allows us to ask whether two ensembles have a \textbf{common component}. That is, then can be seen as different statistical mixture that have some part in common. Two ensemble are separate, noted $\ens[a] \separate \ens[b]$, if they do not have a common component. This allows to characterize the space in terms of statistical decomposition, and see how this differs from other decompositions.

@@

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined \textbf{entropy}. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following:
\begin{enumerate}
	\item strictly concave: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item bounded from above: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{The expression for the Shannon entropy is actually derived instead of imposed axiomatically.} We define two ensembles to be \textbf{orthogonal} when they maximally increase the entropy during mixing. This recovers the standard notion of orthogonality: two classical ensembles are orthogonal when they have disjoint support, and therefore $\int_M \rho_1 \rho_2 d\mu = $; two quantum ensembles are orthogonal when they are defined on orthogonal subspaces, and therefore $\tr(\rho_1 \rho_2) = 0$.

@@

Another crucial feature is that two ensembles maximize the entropy increase if and only if they are orthogonal: two classical ensembles maximize the entropy increase when they have disjoint support, and therefore $\int_M \rho_1 \rho_2 d\mu = 0$; two quantum ensembles when they are defined on orthogonal subspaces, and therefore $\tr(\rho_1 \rho_2) = 0$. This corresponds to the case when the two ensemble are made of instances that are mutually exclusive. Therefore we define two ensembles to be \textbf{orthogonal}, noted $\ens[a] \ortho \ens[b]$, if their mixture leads to a maximal entropy increase. The following additional axiom is valid for orthogonal ensembles
\begin{enumerate}
	\setcounter{enumi}{2}
	\item mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{enumerate}
as mixing elements distinguishable from those of ensemble $\ens[a]$, still gives elements distinguishable from ensemble $\ens[a]$.

It can be shows that orthogonal ensembles are separate, but the converse is not true. In quantum mechanics, for example, all pure states are separate (they cannot be decomposed and therefore cannot have components in common) but not all are orthogonal. In classical spaces, however, all separate ensembles are also orthogonal. Understanding the relationships between these two properties, then, is key in understanding the difference between classical and quantum mechanics, and what can happen in other physical theories.

\begin{tcolorbox}[colback=white, colframe=black]
	Existence of entropy $\Rightarrow$ Strictly concave function whose upper bound defines orthogonality. Statistical mixtures preserve orthogonality.
\end{tcolorbox}

@@

Before introducing the actual measures, we want to note that there is a delicate balance between these structures which is important at both a conceptual level and at a more mathematical level. Part of the difficulty is to recognize which standard modes of understanding the problem should be kept, as the they generalize nicely, and which should be discarded. Let us go through some key examples.

Those that focus on the convex/linear structure may be tempted to close the space under infinite convex/linear combinations. It would be very convenient mathematically, and, in fact, this is what happens in Hilbert spaces. However, this closure leads yo physically ill-defined systems, such as a particle with an infinite average position or undefined energy.(CITE Hilbert) It is the connection to experimental verification, the topology, that will tell us which limits are allowed and which are not. Therefore, the lack of infinite convex is a crucial feature of the space, and convex sets in ensemble spaces are to be understood as closed convex sets, so that closure on the topology includes those, and only those, infinite mixtures that are physically meaningful.

If we focus on probability, we are tempted to extend the intuition from classical mechanics and it is useful to understand exactly why and where it fails. On a convex structure we can define the extreme points, those that cannot be further decomposed. Therefore it is tempting to define \textbf{pure states} as those ensembles that cannot be further decomposed into other ensembles. For example, the pure states for a six-sided die are the ensembles perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i=1}^{6} p_i \ens_i = 1$ of those pure states $\ens_i$. This is exactly how classical discrete spaces work and much of our intuition for probability comes from there.

This approach, however, already becomes problematic in classical mechanics. Since we can take uniform distributions over any region of phase space, we can imagine to take a sequence of more and more refined distributions around a value of position and momentum. Therefore we may be tempted to say that the points of phase space represent perfectly prepared systems, and every other probability is a distribution on these space. This is physically unsound. First of all, we cannot consistently prepare position and momentum with infinite precision, therefore those ensembles do not exist even exist as an idealization. Secondly, the entropy of those ensembles, if they existed, would be minus infinity, which is a problem for any thermodynamic treatment. To make this intuition consistent with the math would require each point in phase space to be topologically isolated, which would extend the $\sigma$-algebra to the power set of $\mathbb{R}^n$, which means there would be no consistent way to define a volume. The fact that the math breaks down is simply a symptom that our physical problem is ill-specified.

In quantum mechanics, things are even worse. Pure states are indeed extreme points of the convex set. However, we also have discrete spectra. Here the temptation is, like we ``did'' in classical mechanics, to consider wave-functions all concentrated at a single value of position. While one may think this is the analogue of classical mechanics, it is not. We can, in fact, take a wave-function that is uniformly distributed on every finite range of position, and we can imagine to shrink that range. However, this is a sequence of pure states. All pure states have zero entropy, and the reduced spread on position will mean an increased spread on momentum. That is, we are not providing a sequence of ensembles at ever greater prevision, like in the classical case. In other words, the eigenstates of position in quantum mechanics, even if we wanted to say they were actual states, they would not be the limit of infinitesimal decomposition (i.e. a limit in the convex structure) but rather the limit of a path over pure states (i.e. a topological limit).

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. convex combinations of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectrum). The two coincide in classical discrete spaces, which are the least physically interesting case as they do not even allow classical particle mechanics. Equivalently, \textbf{mixtures are about preparations, not measurements, and probability distributions are about measurements, not preparations}. In quantum mechanics, an equal mixture of spin up and spin left, for example, does not correspond to a 50\% probability of measuring spin up: the mixing coefficients and the outcome probability are different.

To recap, the topological structure captures experimental verifiability and is responsible for the limits and the connection to measureable quantities. The convex structure captures statistical mixing and is responsible for all linear structures of physical theories. The entropic structure captures the variability within an ensemble and is ultimately responsible for all geometric structures. Since the physical problem is indivisible, there is an interplay across these structure which goes beyond this article and we refer to our larger work.

@@

We have clarified that mixing coefficients are not the same as probabilities of outcomes. Unlike most of the literature on probability, we will concentrate on mixtures and, particularly, on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definition.

@@

In the classical case, the fraction capacity reduces to a probability measure. Note that a probability satisfies all the properties of a fraction capacity, but it also satisfy on additional property: additivity over disjoint sets. That is, if $A \cup B = \emptyset$ then $\mu(A \cup B) = \mu(A) + \mu(B)$. Adding finite additivity over disjoint sets is enough as continuity will extended the property to additivity over countably many disjoint sets. Conversely, having countable additivity over disjoint sets implies sub-additivity and continuity from below and above.

While additivity over disjoint set holds in classical probability, it does not hold in quantum mechanics

 This is not true in general, as we can see in a simple example from quantum mechanics. Let $\Ens$ be the ensemble space of a qubit, that is the space of density matrices of a two dimensional Hilbert space. Geometrically, this is the convex set represented by a three dimensional ball. Let $\ens_{\psi} = |\psi\>\<\psi|$ be a pure state and let $\ens_{\phi} = |\phi\>\<\phi|$ be its orthogonal. Geometrically, these are two opposite points on the surface of the ball. The maximally mixed state can be expressed as $\ens = \frac{1}{2} \ens_0 + \frac{1}{2} \ens_1$. This is the center point of the ball and is in fact the midpoint between two opposite points. Since this is valid for any point, we have $\frcap_{\ens}(\{\ens_{\psi}\})=\frac{1}{2}$ for any $\ens_{\psi} \in \Ens$ and $\frcap_{\ens}(\{\ens_{\psi}, \ens_{\phi}\})=1$ for any pair of orthogonal pure states. If $A$ is a set of two orthogonal pure states, and $B$ is state different from the two, we have $\frcap_{\ens}(A \cup B) = 1 < \frcap_{\ens}(A) + \frcap_{\ens}(B) = \frac{3}{2}$.
